%!TEX root = manual.tex
%
\chapter{Document history}
\label{ch:history}

\asdl{} and \asdlgen{} (originally named \texttt{asdlGen}) were developed as part
of the National Compiler Infrastructure Project at Princeton University in the
late 1990's~\cite{usenix:zephyr-asdl}.
The original version of this manual was written by Dan Wang as part of that project.
As part of reimplementing and modernizing \asdl{}, John Reppy converted the original
manual to \LaTeX{} and did some light editing (this version of the old manual
can be found in the \texttt{doc/manual-2.0} directory).
The original implementation of \texttt{asdlGen} can be
found at \url{http://asdl.sourceforge.net}.

Here is a history of changes to \asdl{}, \asdlgen{}, and this manual.
The changes are indexed by SML/NJ release numbers.

\begin{description}
  \item[SML/NJ 2021.1]
    \mbox{}\\[0.5em]
    Reorganized the \sml{} library to separate input and output into separate
    structures.  For example, the ``\cd{ASDLFilePickle}'' structure is now
    implemented as the pair of structures: ``\cd{ASDLFileReadPickle}'' for input
    and ``\cd{ASDLFileWritePickle}'' for output.  The structure ``\cd{ASDLFilePickle}''
    is provided as a combined structure that provides support for both input
    and output.

  \item[SML/NJ 110.99]
    \mbox{}\\[0.5em]
    Fixed various inconsistencies between the implementation (both \Cplusplus{} and
    \sml{}) and the documented representation.

  \item[SML/NJ 110.98]
    \mbox{}\\[0.5em]
    Initial support for the sharing type operator has been added, but it is incomplete.

    The semantics of the \cd{suppress} view item has been changed to allow more
    fine-grained control.  Instead of a boolean flag, it now takes a list of
    aspects to suppress.

    General improvements to the documentation.

  \item[SML/NJ 110.86]
    \mbox{}\\[0.5em]
    The interface to the file and memory picklers was unified: the ``\cd{encode}''
    was changed to ``\cd{write},'' ``\cd{decode}'' was changed to
    ``\cd{read}'' in the memory pickler, and the read operations were made
    imperative.
    Also added support for writing to an S-Expression textual format and improved
    the documentation.

    Changed the ``\texttt{--pickler}'' option to ``\texttt{--gen}'' and made it take
    a list of targets.

  \item[SML/NJ 110.84]
    \mbox{}\\[0.5em]
    First release of ASDL 3.0; see \secref{sec:changes} for differences from previous
    version.
\end{description}%
